### 2019-07-07 thesis-template/msc-thesis.tex
update

Tensorfelder werden meistens in Verbindung mit mechanischen Spanunngsverteilungen in 2D/3D-Gittern gebracht (vgl. Cauchy-Stress Tensor), haben aber auch andere praktische Bedeutungen in der Physik.
Mithilfe von globalen Beleuchtungsmodellen/-techniken wird eine neue Methode entwickelt, um Tensorfelder zu visualisieren, die aber zusätzlich auch verwendet werden kann um die Lichtausbreitung in einer topologischen Szene zu beschreiben. Als Grundlage leiten wir ein einfaches Lichtausbreitungsschema für kartesische Gitter ab, dass die Prinzipien der Ausbreitungsdämpfung und Energieerhaltung beachtet und Licht-verteilung/en für gegebene Lichtquellenrichtung/en und -positionen bis zur Konvergenz approximieren kann. Als unterliegendes Modell werden die Transmissionsprofile innerhalb diesem Gitter als Kristallfiberstruktur angenommen (vgl. Edelsteine: Katzenaugeneffekt beim Tigerauge). Die folgende Aufgabe ist es, anisotrope Fiberstrukturen mit der Orientierung und den Anisotropiemaßen des unterliegenden Tensorfelds zu modellieren. Dafür erstellen wir per Hauptachsentransformation (PCA) ein Eigensystem für jede Zelle des Tensorfelds und modellieren eine zugehörige Ellipsengleichung als (Transfer-/Wichtungs-) Transmissionsprofil. Folglich kann die resultierende Lichtverteilung als 2D Skalarfeld oder Polarplot für Richtungsinformation visualisiert/repräsentiert werden. Wir messen Impulsantworten des Tensorfelds mit Delta-Pulsen an jeder Position und in jede Richtung um eine uniform gesamplte Map der globalen Lichtverteilungen zu erhalten. Diese Map wird als ein globales Energieflussfeld betrachtet. Wir lassen uns von der Technik FTLE aus dem Bereich Particle Tracing der Vector Field Visualization inspirieren, indem wir den Gradient des Flussfelds der resultierenden Lichtverteilungen analysieren und visualisieren. Ähnlich wie FTLE in Vektorfeldern Gräte/Kämme/LCS detektiert, erfasst unsere Methode dieselben Strukturen, anziehende/abstoßende/sattelnde Tensorfeldlinien die häufig Schlüsselstrukturen in der Natur repräsentieren, in Tensorfeldern. Wir führen diese Größe als Global Illumination Gradient ein, der einen FTLE-verwandten Ansatz darstellt, um LCS in Tensorfeldern durch die Analyse des globalen, durch einen Imprint (Gravur in kristallinen Fiberstrukturen) gerichteten Lichtfluss, zu visualisieren.
\newpage

\chapter*{Abstract}
%\input{abstract}
Many problems in science and engineering require
tensor field representations. Tensor fields, are commonly associated with stress distributions (cf.~Cauchy-stress tensor) in 2D/3D grids, but also have many other meanings in various fields. We present the concept of applying techniques from the field of global illumination to tensor field analysis. Thereby, we interpret the tensors as a conductivity property for light transport. As a basis, we derive a simple light propagation scheme for Cartesian grids, which satisfies propagation attenuation and energy-conservation principles, and is able to approximate light distributions for given light source position(s) and direction(s). The transmission profiles within this grid are considered as \enquote{crystal fiber structures} as an underlying physical model (e.g.: gemstones: tiger's eye's cat's eye effect). The consequent task is to interpret anisotropic fiber structures with the orientation and anisotropy measures of the underlying tensor field in the grid. For this, we derive an eigenbasis by principal component analysis for every cell of the tensor field and form an ellipsoid equation as transmission (conductivity) function for the transmitted light profiles. Hence, the resulting light distribution can be visualized via 2D scalar field color coding or polar plots for directional information. On this basis, we measure impulse responses of the tensor field with Dirac pulses as light sources at any (sampled) position and in any direction to generate a uniform sampled map of the global illumination distributions. This map is considered as a global illumination energy flow field. We gain inspiration by vector field particle tracing's FTLE approach for analyzing the gradient of the flow field, formed by the resulting light distributions. Similar to how the FTLE detects ridges/Lagrangian coherent structures in vector fields, our approach captures these structures likewise, namely regions revealing attracting/repelling/saddling tensor field lines which frequently represent key structures in nature, in tensor fields. We denote this entity as the light transport gradient (LTG), which states an FTLE-related approach for visualizing LCS in tensor fields through analysis of the global light transport in the grid, directed by imprints in crystelline fiber structures modeled by the tensor field. We evaluate our approaches for plausibility and comparability within an extensive evaluation and demonstrate its performance on several examples of both synthetic and real datasets.
\newpage

@@

\section{Motivation}
\label{sec:motiv}
Tensor field visualization is gaining importance as a relevant tool for the analysis of fluid and solid mechanics, where tensor representations are crucially needed. This is always the case, when a directional distribution is needed for each single point in space. Tensors occur most commonly in medical, scientific and engineering applications, making them a frequently used mathematical entity in science. In general, a tensor is a order-$o$ generalization of scalar, vector and matrix and appears, e.g., as the Jacobian matrix (gradient in vector fields), or as stress tensor in solid and fluid continuum mechanics. Basically, tensors describe compressions, tensions, rotations and volume changes in both solid and fluid material, as illustrated in Fig.~\ref{ToS}. While most techniques focus on symmetric tensor field visualization (like glyphs and tensor field lines), some other recent works address the problem of asymmetric tensor field visualization. The issue with asymmetric (non-symmetric) or therein included antisymmetric (skew-symmetric) tensors is, that they do not always yield real eigenvalues, which is needed to determine a real eigenbasis to set up glyphs, hyperstreamlines or tensor field lines. Nevertheless, there has been extensive work in this domain as well. Classically, the issue with asymmetric tensor fields is, that they do not yield real eigenvalues in general, which is needed to set up glyph representations. In this work, we will use the matrix decomposition obtained from singular value decomposition (SVD) to obtain the half-axes of the PCA principal ellipsoid representing the tensor transformation as suggested by Moler \cite{moler} to state another alternative of processing asymmetric tensors. 2D tensor field visualization is useful for cases where only two-dimensional data is available, such as satellite remote sensing, particle imagery velocimetry and earthquake data \cite{laramee}. In their field of work, geologists and medical scientists are struggling with complicated number arrays which seem almost impossible to interpret in their raw, unprocessed form. Nevertheless, comprehensive tensor field visualization is aiding scientists in their everyday lifes understanding complex data loads, such as tensor fields. To cope with the arising problem of tensor field visualization, one needs to advance and extend respective techniques.

@@

The main purpose of visualization in general, is to reduce or encode some set of numbers in some readable, explorable and more intuitive representation. This implies, e.g., clustering, structuring or projection techniques. When it comes to tensor fields, this comes in handy in, e.g., medical sciences such as in DT-MRI (diffusion tensor - medical resonance imaging) \cite{hlawatsch}, when measuring water molecule diffusion characteristics within tissue or in simulated or measured stress tensor fields from solid and fluid mechanics. In these named domains of science the problem arises, that tensor fields can be generated or measured with physical/mathematical background, but not be regarded intuitively. Necessity begets ingenuity as always, which lets visualization methods and systems emerge in this context. Concurrently, the problem setting of this work encomprises the visualization of tensor fields, i.e., mostly 2D/3D-grids, commonly consisting of second-order tensors (cf.~Cauchy stress tensor). In general, these grids span a spatial domain with sample positions indicated by vertices or cell centers in the dual grid, depending on the mode of interpretation. Tensor fields feature a sampling grid, consistent of tensors, which is spaced uniformly in the commonly used regular grid. A Cartesian grid, which we will utilize in this context, is a grid normalized to a certain unit length, as a special case of a regular grid. That is, each cell is formed as a unit cube in relation to the normalizing unit length. We aim to use global illumination methods to model light transport, directed by tensor fields, within such type of grids, which we consider as light propagation volumes (LPVs) \cite{dachsbacher}. Congruently, the most basic objective of this work is to design a simple light propagation scheme following the most basic but crucial physical principles and laws, like energy conservation and propagation attenuation, for tensor field visualization. Of course, the propagation scheme will need some kind of stimulus which can be defined through symbolic functions (e.g.: $r(\omega)=1$) by the user which we sample (discretize) as a primary step. This way, the approach will operate as a light transport visualization technique for light distributions given in polar coordinates, provided the user set an initial profile into the grid. Furthermore, this propagation scheme should either be used to visualize the light transport paths, defined by the anisotropy characteristics of tensor fields and tensor magnitudes of a given tensor field in a $2D$ grid. Or, as a secondary purpose, be used to visualize the light transport in a defined, given geometrical scene in a $2D$ grid. However, the tensor field needs to be preprocessed somehow to generate a footprint map of transmission profiles defined by the tensor field defined by the tensor field's anisotropy characteristics, which we will handle through matrix decompositions. For this purpose, we will let the transmission profiles act on our simple light propagation scheme to have a kind of tensor controlled directive effect on the light transport. The transmission profiles should modulate the outgoing intensities in a sense of redirection or more appropriately, redistribution. This will yield an approach, capable of computing approximated final light distributions for a given light source configuration inside of a $2D$ lattice. As this process requires a lot of user-interaction and is not to be concerned as generalizing to all possible scenarios. In the sense of automization and optimzation, another consequent aim is to implement an approach borrowed from vector-field visualization to reveal structures, which are hard to grasp by ordinary methods. We construct an FTLE-related approach in tensor field visualization (both symmetric and asymmetric) for detecting ridges representing LCS (Lagrangian Coherent Structures). These structures can be classified (identified) as repelling, attracting and saddling (forming a saddle) local features, which frequently represent key structures in nature, such as moraines in glaciers or vortices in flow fields. Commonly, in vector field visualization, LCS separate regions of different flow behavior, but in our case they separate regions/subdomains of varying directional distributions (manifesting in diverging/converging tensor field lines leading to compressions, tensions, roations or varying tensor magnitude causing volume changes) represented by tensors, as we use our tensor-controlled light transport visualization (propagation scheme) as a basis or framework to compute the FTLE field. Consequently, we propose another method to compute an FTLE-like field on tensor fields. The FTLE field responds most where tensor field lines / hyperstreamlines converge or diverge, whereas one is the same as the other because of bidirectional stress directions and since we effectively measure the finite-time Lyapunov exponent (FTLE) with both forward and backward integration time (stimulus directions, in our case), we will detect both cases. Both invented methods, will give new insights into the topology and anisotropy of tensor fields and provide new tools for the structural analysis. The greater goal of this work is therefore the encoding of a tensor field into certain abstracted representations and the interpretation of meaningful results evaluating the proposed techniques for possible uses. The designed approach could, e.g., be applied in medicinal sciences, engineering and physics to reveal structures which are hard to grasp through interpreting simpler and sometimes confusing glyph and tensor field line visualizations. A classical example for tensor field visualization is plotting hyperstreamlines following major/minor eigenvectors, as depicted in Fig.~\ref{streamlines}, which features a degenerate point (purple), a wedge (red) and a trisector (green).

@@

For the rest of this work, we will give a quick overview on the structure and outline. Next, in Chap.~\ref{chap:grundlagen}, we will capture important fundamentals like tensors, polar coordinates, PCA in the context of singular value decomposition (generalization of eigenvalue decomposition), FTLE and GT (ground truth) datasets. Also we will discuss the most relevant related work in the domain of tensor field visualization and also in the domain of global illumination (indirect illumination rendering). Subsequently, in Chap.~\ref{chap:meth}, we will propose the main contributions and methods of this work and document the results and evaluation in Chap.~\ref{chap:eval}. At last, a compact summary, conclusion and outlook will be given in Chap.~\ref{chap:concl}.\\

@@

There has been extensive recent work on symmetric tensors but in comparison little on asymmetric ones. Since we are not able to cover the whole scope in this work, we will focus on the most relevant works on tensor field visualization. We will talk about the publications from Hlawatsch et al. \cite{hlawatsch} and Zirr et al. \cite{zirr} in particular, since they constitute thematically close related works. But first, we will cover related works on global illumination methods, as these form a basis or prerequisite for our work. Please note, that they do not focus on tensor field visualization at all. Remember, that we will inventively employ global illumination methods for tensor field visualization as an innovative step and that we want to cover previous works on our base approach, as well.

\section{Global Illumination Methods}
Since we use a specific category of global illumination methods as a basis, we decide to only discuss lattice based methods which work, corresponding to their name, on a grid. Remark, that Global Illumination methods classically focus on realistic indirect lighting. Thus, they aim on producing a most realistic image, what in our case turns out to be a most informative image in correspondence. The most early work in this field is propably the Discrete Ordinates Method (DOM) by Chandrasekhar \cite{Chandrasekhar}, which is used specifically for computing radiative transfer in participating media (volume rendering). The radiative transfer equation is discretized spatially and in the angular domain. The radiance distribution is stored in a grid and light is distributed and exchanged between neighboring volume elements which reduces the overall effort to local and independent computations. Furthermore, there has been more recent work on describing the light propagation as a diffusion process within a simple photon transport model based on the Lattice-Boltzmann method \cite{geist}. DOMs were recently revisited by Fattal \cite{fattal}. This work addresses artifacts stemming from spatial discretization, which are called \enquote{ray effects}, and from successive interpolation called \enquote{light smearing}. These methods are considered to be too slow to \enquote{run at interactive framerates} by Dachsbacher et al. \cite{dachsbacher}, though \enquote{efficient and potentially highly parallel}. Our work is inspired by Lattice-Boltzmann methods (volume grid radiosity) and the lattice-based discrete ordinate methods concerning the light propagation scheme.

\section{Symmetric Tensor Field Visualization}
Zheng \& and Pang \cite{pang&zheng} et al. proposed a texture-based visualization approach called Hyper LIC, which extends the concept of LIC to symmetric tensor fields by using an anisotropic 2D-filter kernel oriented along the major/minor eigenvectors. The concept of following the major/minor eigenvector along tensor field lines was invented by Vilanova et al. \cite{vilanova}. Tensorlines \cite{weinstein} introduce a kind of artificial inertia (running average), making them resistent to noise. The notion of tensor field topology (critical, degenerate points, separatrices) and the concept of hyperstreamlines was first introduced by Delmarcelle and Hesselink \cite{delmarcelle&hesselink}. The topology in $3D$ tensor fields was first analyzed by Zheng et al. \cite{zheng}. The placement of tensor field lines as hyperstreamlines for glyph packing was recently improved by Spencer et al. \cite{spencer}. A good overview of seeding strategies concerning glyphs is given by McLoughlin et al. \cite{mcloughlin}. Feng et al. \cite{feng} used Voronoi Tesselation for placing the glyphs. Superquadric glyphs have been proposed by Schultz and Kindlmann \cite{schultz&kindlmann} for general (non-positive definite) tensors. De Leeuw and van Wijk \cite{deleeuw&vanwijk} visualize the partial derivative gradient, the Jacobian of the tensor field, to sense the local properties of the field. A diffusion tensor can also be visualized by box \cite{makris}, ellipsoid \cite{pierpaoli}, composite \cite{westin2} or superquadric \cite{kindlmann} glyphs. There is also a set of scalar measures available in symmetric tensor field visualization such as: mean diffusivity, fractional anisotropy \cite{basser} and anisotropy coefficients \cite{westin}. These coefficients can also be visualized by direct volume rendering \cite{kindlmann2}. Hlawatsch et al. \cite{hlawatsch} obtain a coordinate (flow) map from tensor field lines following major/minor eigenvectors (fibre trajectories) using the maximum eigenvalue of the Cauchy-Green deformation tensor to compute the FTLE-field in homogeneous regions. In a similar manner, we aim to use light transport techniques to obtain a final light distribution (flow map) through successive light propagation cycles directed (in form of transmission profiles obtained from ellipsoid glyphs) by the tensor field. DT-MRI diffusion tensor imaging can help to reveal the functional \cite{denis} or topological \cite{wakana} structure of the brain and muscles by analysing celebral and muscle tissue tractography, which yields a tensor field representing the anisotropy characteristcs of the brain's white matter. An overview about DT-MRI visualization techniques is given by \cite{vilanova}. Tricoche et al. \cite{xavier} use invariant manifolds to extract topological features from tensor fields. Other feature extraction-based approaches extract general features as: edges \cite{hancock}, tensor shape\/ orientation \cite{gordon}, interfaces \cite{donnel}, creases \cite{tricochet}.
\\
\section{Asymmetric Tensor Field Visualization}
Zheng and Pang \cite{pang&zheng} proposed the concept of dual eigenvectors and Zhang et al. \cite{zhang} extended it to pseudo-eigenvectors and introduced the eigenvector/eigenvalue manifolds to visualize eigenvectors in the complex domain. Laramee et al. \cite{laramee} focused on the efficient implementation and visualization of these structures and provided an interactive visualization system for asymmetric tensors applicable in fluid and solid dynamics. The concept of tensor magnitude has been introduced by \cite{laramee} for means of physical interpretation. They also proposed an efficient glyph and hyperstreamline hybrid approach, which made dynamic interaction in real-time in $2D$ tensor fields feasible. Palacios et al. \cite{palacios} extracted isosurfaces of tensor magnitude, mode and isotropy index.

>>>>

Tensorfelder werden häufig in Verbindung mit mechanischen Spanunngsverteilungen in 2D/3D-Gittern gebracht (vgl. Cauchy-Stress Tensor), haben aber auch andere praktische Bedeutungen in der Physik.
Mithilfe von globalen Beleuchtungsmodellen/-techniken wird eine neue Methode entwickelt, um Tensorfelder zu visualisieren, die aber alternativ auch verwendet werden kann um die Lichtausbreitung in anisotropen Medien zu beschreiben. Als Grundlage leiten wir ein einfaches Lichtausbreitungsschema für kartesische Gitter ab, welches die Prinzipien der Ausbreitungsdämpfung und Energieerhaltung beachtet und Lichtverteilung/en für gegebene Lichtquellenrichtung/en und -positionen approximieren kann. Als unterliegendes Modell werden die Transmissionsprofile innerhalb diesem Gitter als Kristallfiberstruktur angenommen (vgl. Edelsteine: Katzenaugeneffekt beim Tigerauge). Die folgende Aufgabe ist es, anisotrope Fiberstrukturen mit der Orientierung und den Anisotropiemaßen des unterliegenden Tensorfelds zu interpretieren. Dafür erstellen wir per Hauptachsentransformation ein Eigensystem für jede Zelle des Tensorfelds und modellieren eine zugehörige Ellipsengleichung als (Wichtungs-) Transmissionsprofil. Dabei werden die physikalischen Eigenschaften des Tensorfelds (Orientierung/Anisotropie/Amplitude) als Lichtleitfähigkeit interpretiert und die Lichtverteilung berechnet. Folglich kann die resultierende Lichtverteilung als 2D Skalarfeld oder Polarplot für zusätzliche Richtungsinformation repräsentiert und visualisiert werden. Wir messen Impulsantworten des Lichtausbreitungs-Prozesses moduliert mit den anisotropen Charakteristika des Tensorfelds, indem Delta-Pulse an jeder Position und in jede Richtung als Stimuli verwendet werden. Wir lassen uns von der Methode FTLE aus dem Bereich Particle Tracing der Vector Field Visualization inspirieren, indem wir den Gradient des Flussfelds der resultierenden Lichtverteilungen analysieren und visualisieren. Ähnlich wie FTLE in Vektorfeldern Gräte/Kämme/LCS detektiert, erfasst unsere Methode dieselben Strukturen, nämlich anziehende/abstoßende/sattelnde Tensorfeldlinien die häufig Schlüsselstrukturen in der Natur repräsentieren, in Tensorfeldern. Wir führen diese Größe als Light Transport Gradient ein, der einen FTLE-verwandten Ansatz darstellt, um LCS in Tensorfeldern durch die Analyse des globalen, durch einen Imprint (Gravur in kristallinen Fiberstrukturen) gerichteten Lichtfluss, zu visualisieren.
\newpage

\chapter*{Abstract}
%\input{abstract}
Many problems in science and engineering require
tensor field representations. Tensor fields, are commonly associated with stress distributions in 2D/3D grids, but also have many other meanings in various fields. By means of global illumination techniques, we motivate a new method to visualize tensor fields,
which is also capable of visualizing the light transport (propagation) in anisotropic media. We present the concept of applying techniques from the field of global illumination to tensor field analysis. Thereby, we interpret the tensors as a conductivity property for light transport. As a basis, we derive a simple light propagation scheme for Cartesian grids, which satisfies propagation attenuation and energy-conservation principles, and is able to approximate light distributions for given light source position(s) and direction(s). The transmission profiles within this grid are considered as \enquote{crystal fiber structures} as an underlying physical model (e.g.: gemstones: tiger's eye's cat's eye effect). The consequent task is to interpret anisotropic fiber structures with the orientation and anisotropy measures of the underlying tensor field in the grid. For this, we derive an eigenbasis by principal component analysis for every cell of the tensor field and form an ellipsoid equation as transmission (conductivity) function for the transmitted light profiles. Hence, the resulting light distribution can be visualized via 2D scalar field color coding or polar plots for directional information. On this basis, we measure impulse responses of the tensor field with Dirac pulses as light sources at any (sampled) position and in any direction to generate a uniform sampled map of the global illumination distributions. This map is considered as a global illumination energy flow field. We gain inspiration by vector field particle tracing's FTLE approach for analyzing the gradient of the flow field, formed by the resulting light distributions. Similar to how the FTLE detects ridges/Lagrangian coherent structures in vector fields, our approach captures these structures likewise, namely regions revealing attracting/repelling/saddling tensor field lines which frequently represent key structures in nature, in tensor fields. We denote this entity as the light transport gradient (LTG), which states an FTLE-related approach for visualizing LCS in tensor fields through analysis of the global light transport in the grid. We evaluate our approaches for plausibility and comparability within an extensive evaluation and demonstrate its performance on several examples of both synthetic and real datasets.
\newpage

@@

\section{Motivation}
\label{sec:motiv}
Tensor field visualization is gaining importance as a relevant tool for the analysis of fluid and solid mechanics, where tensor representations are crucially needed. This is always the case, when a directional distribution is needed for each single point in space. Tensors occur most commonly in medical, scientific and engineering applications, making them a frequently used mathematical entity in science. In general, a tensor is a order-$o$ generalization of scalar, vector and matrix and appears, e.g., as the Jacobian matrix (gradient in vector fields), or as stress tensor in solid and fluid continuum mechanics. Basically, tensors describe compressions, tensions, rotations and volume changes in both solid and fluid material, as illustrated in Fig.~\ref{ToS}. While most techniques focus on symmetric tensor field visualization (like glyphs and tensor field lines), some other recent works address the problem of asymmetric tensor field visualization. The issue with asymmetric (non-symmetric) or therein included antisymmetric (skew-symmetric) tensors is, that they do not always yield real eigenvalues, which is needed to determine a real eigenbasis to set up glyphs, hyperstreamlines or tensor field lines. Nevertheless, there has been extensive work in this domain as well. Classically, the issue with asymmetric tensor fields is, that they do not yield real eigenvalues in general, which is needed to set up glyph representations. In this work, we will use the matrix decomposition obtained from singular value decomposition (SVD) to obtain the half-axes of the PCA (principal component analyis) principal ellipsoid representing the tensor transformation as suggested by Moler \cite{moler} to state another alternative of processing asymmetric tensors. 2D tensor field visualization is useful for cases where only two-dimensional data is available, such as satellite remote sensing, particle imagery velocimetry and earthquake data \cite{laramee}. In their field of work, geologists and medical scientists are struggling with complicated number arrays which seem almost impossible to interpret in their raw, unprocessed form. Nevertheless, comprehensive tensor field visualization is aiding scientists in their everyday lifes understanding complex data loads, such as tensor fields. To cope with the arising problem of tensor field visualization, one needs to advance and extend respective techniques.

@@

The main purpose of visualization in general, is to reduce or encode some set of numbers in some readable, explorable and more intuitive representation. This implies, e.g., clustering, structuring or projection techniques. When it comes to tensor fields, this comes in handy in, e.g., medical sciences such as in DT-MRI (diffusion tensor - medical resonance imaging) \cite{hlawatsch}, when measuring water molecule diffusion characteristics within tissue, or furthermore in simulated or measured stress tensor fields from solid and fluid mechanics. In these named domains of science the problem arises, that tensor fields can be generated or measured with physical/mathematical background, but not be regarded intuitively. Necessity begets ingenuity as always, which lets visualization methods and systems emerge in this context. Concurrently, the problem setting of this work encomprises the visualization of tensor fields, i.e., mostly 2D/3D-grids, commonly consisting of second-order tensors (cf.~Cauchy stress tensor). In general, these grids span a spatial domain with sample positions indicated by vertices or cell centers in the dual grid, depending on the mode of interpretation. Tensor fields feature a sampling grid, consistent of tensors, which is spaced uniformly in the commonly used regular grid. A Cartesian grid, which we will utilize in this context, is a grid normalized to a certain unit length, as a special case of a regular grid. That is, each cell is formed as a unit cube (square) in relation to the normalizing unit length. We aim to use global illumination methods to model light transport, directed by tensor fields, within such type of grids, which we consider as light propagation volumes (LPVs) \cite{dachsbacher}. Congruently, the most basic objective of this work is to design a simple light propagation scheme following the most basic but crucial physical principles and laws, like energy conservation and propagation attenuation, eventually to visualize tensor fields. Of course, the propagation scheme will need some kind of stimulus which can be defined through symbolic functions (e.g., $r(\omega)=1$) by the user which we sample (discretize) as a primary step. This way, the approach will operate as a light transport visualization technique for light distributions given in polar coordinates, provided the user set an initial profile into the grid. Furthermore, this propagation scheme should either be used to visualize the light transport paths, defined by the anisotropy characteristics of tensor fields and tensor magnitudes of a given tensor field in a 2D grid. Or, as a secondary purpose, be used to visualize the light transport in a defined, given geometrical scene in a $2D$ grid. However, the tensor field needs to be preprocessed somehow to generate a footprint map of transmission profiles defined by the tensor field's anisotropy characteristics, which we will handle through PCA yielding a principal component ellipsoid. For this purpose, we will let the transmission profiles act on our simple light propagation scheme to have a kind of  directive effect on the light transport controlled by the tensor's anisotropy characteristics. The transmission profiles should modulate the outgoing intensities in a sense of redirection or more appropriately, redistribution. This will yield an approach, capable of computing approximated final light distributions for a given light source configuration inside of a 2D lattice. As this process requires a lot of user-interaction and is not to be concerned as generalizing to all possible scenarios, we aim to do something more elaborate on the basis of the light propagation scheme. We place light sources in form of Dirac pulses at any position and pointing in any direction represented in the grid, which represents a flow map, which we will analyze for its gradient to reveal the topolgical structure of the tensor field. In the sense of automization and optimization, another consequent aim is to implement an approach borrowed from vector-field visualization to reveal structures, which are hard to grasp by ordinary methods. We construct an FTLE-related approach in tensor field visualization (both symmetric and asymmetric) for detecting ridges representing LCS (Lagrangian Coherent Structures). These structures can be classified (identified) as repelling, attracting and saddling (i.e., forming a saddle) local features, which frequently represent key structures in nature, such as moraines in glaciers or vortices in flow fields. Commonly, in vector field visualization, LCS separate regions of different flow behavior, but in our case they separate regions, or in other words subdomains, of varying directional distributions (manifesting in diverging/converging tensor field lines leading to compressions, tensions, roations or varying tensor magnitude causing volume changes) represented by tensors, as we use our tensor-controlled light transport visualization (propagation scheme) as a basis or framework to compute the FTLE field. Consequently, we propose another method to compute an FTLE-like field on tensor fields. The FTLE field responds most where tensor field lines / hyperstreamlines converge or diverge, whereas one is the same as the other conceptually, because of bidirectional stress directions. Since we effectively measure the finite-time Lyapunov exponent (FTLE) with both forward and backward integration time (stimulus directions, in our case), we will detect both cases dependent on our initial light source direction. Both invented methods, will give new insights into the topology and anisotropy of tensor fields and provide new tools for the structural analysis. The greater goal of this work is therefore the encoding of a tensor field into certain abstracted representations and the interpretation of meaningful results evaluating the proposed techniques for possible uses. The designed approach could, e.g., be applied in medicinal sciences, engineering and physics to reveal structures which are hard to grasp through interpreting simpler and sometimes confusing glyph and tensor field line visualizations. A classical example for tensor field visualization is plotting hyperstreamlines following major/minor eigenvectors, as depicted in Fig.~\ref{streamlines}, which features a structurally unstable point (purple), a wedge (red) and a trisector (green).

@@

For the rest of this work, we will give a quick overview on the structure and outline. Next, in Chap.~\ref{chap:grundlagen}, we will capture important fundamentals like tensors, polar coordinates, PCA in the context of singular value decomposition (generalization of eigenvalue decomposition), FTLE and GT (ground truth) datasets. Furthermore, we will discuss the most relevant related work in the domain of tensor field visualization and also in the domain of global illumination (indirect illumination rendering), since we use these methods as template models, as well. Subsequently, in Chap.~\ref{chap:meth}, we will propose the main contributions and methods of this work and document the results and evaluation in Chap.~\ref{chap:eval}. At last, a compact summary, conclusion and outlook will be given in Chap.~\ref{chap:concl}.\\

@@

There has been extensive recent work on symmetric tensors but in comparison little on asymmetric ones. Since we are not able to cover the whole scope in this work, we will focus on the most relevant works on global illumination and tensor field visualization. We will talk about the publications from Hlawatsch et al. \cite{hlawatsch} and Zirr et al. \cite{zirr} in particular, since they constitute thematically close related works. But first, we will cover related works on global illumination methods, as these form a basis or prerequisite for our work. Please note, that they do not focus on tensor field visualization at all. Remember, that we will inventively employ global illumination methods for tensor field visualization as an innovative step and that we want to cover previous works on our base approach, as well.

\section{Global Illumination Methods}
Since we use a specific category of global illumination methods as a basis, we decide to only discuss lattice based methods which work, corresponding to their name, on a grid. Remark, that Global Illumination methods classically focus on realistic indirect lighting. Thus, they aim on producing a most realistic image, what in our case turns out to be a most informative image in correspondence. The most early work in this field is propably the Discrete Ordinates Method (DOM) by Chandrasekhar \cite{Chandrasekhar}, which is used specifically for computing radiative transfer in participating media (for volume rendering). The radiative transfer equation is discretized spatially and in the angular domain. The radiance distributions are represented in a grid and light is distributed and exchanged between neighboring volume elements which reduces the overall effort to local and independent computations. Furthermore, there has been more recent work on describing the light propagation as a diffusion process within a simple photon transport model, based on the Lattice-Boltzmann method \cite{geist}. DOMs were recently revisited by Fattal \cite{fattal}. This work addresses artifacts emerging from spatial discretization, which are called \enquote{ray effects}, and from successive interpolation called \enquote{light smearing}. These methods are considered to be too slow to \enquote{run at interactive framerates} by Dachsbacher et al. \cite{dachsbacher}, though \enquote{efficient and potentially highly parallel}. Our work is inspired by Lattice-Boltzmann methods (volume grid radiosity) and the lattice-based discrete ordinate methods concerning the light propagation scheme.

\section{Symmetric Tensor Field Visualization}
Zheng \& and Pang \cite{pang&zheng} et al. proposed a texture-based visualization approach called HyperLIC, which extends the concept of LIC to symmetric tensor fields by using an anisotropic 2D-filter kernel oriented along the major or minor eigenvectors. The concept of following the major/minor eigenvector along tensor field lines was first invented by Vilanova et al. \cite{vilanova}. Tensorlines \cite{weinstein} introduce a kind of artificial inertia (running average), making them more resistent to noise. The notion of tensor field topology (critical, degenerate points, separatrices) and the concept of hyperstreamlines was first introduced by Delmarcelle and Hesselink \cite{delmarcelle&hesselink}. The topology in $3D$ tensor fields was first analyzed by Zheng et al. \cite{zheng}. The placement of tensor field lines as hyperstreamlines for glyph packing was recently improved by Spencer et al. \cite{spencer}. A good overview of seeding strategies concerning glyphs is given by McLoughlin et al. \cite{mcloughlin}. Feng et al. \cite{feng} used Voronoi tessellation for placing the glyphs. Superquadric glyphs have been proposed by Schultz and Kindlmann \cite{schultz&kindlmann} for general (non-positive definite) tensors. De Leeuw and van Wijk \cite{deleeuw&vanwijk} visualize the partial derivative gradient, the Jacobian of the tensor field, to sense the local properties of the field. A symmetric tensor can also be visualized by box \cite{makris}, ellipsoid \cite{pierpaoli}, composite \cite{westin2} or superquadric \cite{kindlmann} glyphs. There is also a set of scalar measures available in symmetric tensor field visualization such as: mean diffusivity, fractional anisotropy \cite{basser} and anisotropy coefficients \cite{westin}. These coefficients can also be visualized by direct volume rendering \cite{kindlmann2}. Hlawatsch et al. \cite{hlawatsch} obtain a coordinate (flow) map from tensor field lines following major/minor eigenvectors (fibre trajectories) using the maximum eigenvalue of the Cauchy-Green deformation tensor to compute the FTLE-field in homogeneous regions. In a similar manner, we aim to use Global Illumination methods propagating light, directed by tensor fields, to obtain a final light distribution (flow map) through successive light propagation cycles. The light distributions are redirected by the tensor field employing transmission profiles obtained from ellipsoid glyphs. DT-MRI diffusion tensor imaging can help to reveal the functional \cite{denis} or topological \cite{wakana} structure of the brain and muscles by analysing celebral and muscle tissue tractography, which yields a tensor field representing the anisotropy characteristcs of the brain's white matter. An overview about DT-MRI visualization techniques is given by \cite{vilanova}. Tricoche et al. \cite{xavier} use invariant manifolds to extract topological features from tensor fields. Other feature extraction-based approaches extract general features as: edges \cite{hancock}, tensor shape\/ orientation \cite{gordon}, interfaces \cite{donnel} and creases \cite{tricochet}.
\\
\section{Asymmetric Tensor Field Visualization}
Zheng and Pang \cite{pang&zheng} proposed the concept of dual eigenvectors and Zhang et al. \cite{zhang} extended it to pseudo-eigenvectors and introduced the eigenvector/eigenvalue manifolds to visualize eigenvectors in the complex domain. Laramee et al. \cite{laramee} focused on the efficient implementation and visualization of these structures and provided an interactive visualization system for asymmetric tensors applicable in fluid and solid dynamics. The concept of tensor magnitude has been introduced by \cite{laramee} for means of physical interpretation. They also proposed an efficient glyph and hyperstreamline hybrid approach, which made dynamic interaction in real-time in $2D$ tensor fields feasible. Palacios et al. \cite{palacios} extracted isosurfaces of tensor magnitude, mode and isotropy index.
